|

### 2021-10-03 article/article.Rmd
text edits, article and appendix

---
title:  |
        | Legislature Size and Welfare:
        | Evidence from Brazil
author: 
  - Gabriel Cepaluni[^gabriel]
  - Danilo Freire[^danilo]
  - Umberto Mignozzetti[^umberto]
date: \today
thanks: "We thank Cedric Antunes, Victor Araujo, Eric Arias, Taylor Boas, Natalia Bueno, Andrea Freitas, Saad Gulzar, James Mahon, Lucas Novaes, Nelson Ruiz, Catarina Roman, David Skarbek, and Denis Stukal for their valuable comments. We also thank the participants of seminars and conferences at CEPESP, CUNY, INSPER, IPSA, LACEA, LASA, MPSA, Notre Dame, NYU, PUC Chile, SPSA, and Uppsala. We gratefully acknowledge the financial support of the São Paulo Research Foundation (FAPESP), grant number 2017/07290-5. All materials required to replicate this study are available at <https://github.com/umbertomig/legislature-size-welfare>."
abstract: "What is the effect of the number of legislators on public service provision? Despite the importance of institution designs for young democracies, little is known about the particular features that may help improve welfare. This paper uses a formal model to show how increasing legislature size may affect service provision. I show that changes in bargaining costs depend on whether the additional legislator shares the Executive's party affiliation: More opposition members reduce the Executive's equilibrium public goods provision, while more government-aligned members increase it. I test this theory by exploiting sharp discontinuities in city council size in Brazil. I show that an additional city councilor has a 91% chance of belonging to the mayoral coalition, significantly improving primary school enrollment and infant mortality. To investigate the mechanism, I surveyed former councilors and analyzed 346,553 bills proposed between 2005 and 2008. This paper has implications for the design of representative institutions."
abstractspacing: double
keywords: Legislature Size; Legislative Bargaining; Local Public Goods Provision; Regression Discontinuity Design; Brazil
jelcodes: H41; H80; I31; I38
fontsize: 11pt
margin: 2cm
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
    md_extensions: +raw_attribute
---

@@

[^danilo]: Senior Lecturer, School of Social and Political Sciences, University of Lincoln, <danilofreire@gmail.com>, <https://danilofreire.github.io>.

@@

Legislative institutions are crucial for welfare and service provision in democratic countries. Historically, the division of power and the birth of a representative government placed the legislatures at the center of popular consent. In modern democracies, most countries have assemblies at all levels of government. These institutions affect the information and quality of legislation, check and veto the other political powers, and influence policymaking. Moreover, they also determine taxation, government spending and oversee service provision \citep{Weingastetal1981, weingast1988industrial, north1989constitutions, auriol2012optimal, stasavage2011states, poulsen2018political}.

Regardless of differences in electoral rules, social composition, and procedures, all legislatures share one feature: more than one representative composes them. Legislature size relates to collective action problems, bargaining between political powers, government spending, and representation, but how these components affect welfare remain understudied \citep{crain1979cost, Weingastetal1981, rogers2002free, PrimoSnyder2008, allenrepresentation}. On the one hand, enlarged legislatures may increase government spending and lawmaking, and this could improve some services \citep{Weingastetal1981}. On the other hand, adding a legislator may increase the number of veto players and broaden the collective action problems, hindering service provision \citep{tsebelis2002veto, gottlieb2019countervailing}. Therefore, what is the effect of increasing the legislature size on public service provision?

In this paper, I propose a local legislature size and welfare theory that incorporates two crucial features: Executive leadership and party politics. First, in mayor-council municipalities, mayors tend to be central in the local public service decisions. Mayors have the prerogative in proposing public investments, much as presidents in presidential systems. Second, partisanship is essential at the local level, even in settings where the ideological attachment is weak, such as in most young democracies. Party attachment helps structures political views \citep{samuels2014power}, and local-level elites usually belong to different factions and parties. They fight for power in the municipality by controlling various parties and forming distinct pre-electoral coalitions.\footnote{Elite political competition in the local level has also been linked to violence and conflict in Brazil and elsewhere \citep{hoelscher2015politics, mignozzetti2018predatory}.} In most cities, family ties are substitutes for the ideological component, but the most influential local leaders sort themselves in different political groups \citep{nunes1977coronelismo, ferraz2020political}.

This paper incorporates partisanship and mayoral first-move advantage to a \citet{baron1989bargaining} bargaining model. Mayors make the initial policy proposal that consists of an investment in public goods provision, a decision to extract rents for personal gains, and a vector of side-transfers for councilors. If they fail to secure majoritarian support from councilors, a reversal stage occurs, where the decision leaves the mayor's office, and councilors then propose the public investments. In this mayor-council bargaining, costs come from the side-transfers required to secure a majority among councilors. I show that in a model without partisan concerns, bargaining costs always increase with larger city council size, lowering public goods provision and rents. However, adding partisanship decreases the bargaining costs when the chances of electing councilors aligned with the mayor are sufficiently high. Lower bargaining costs release funds for mayors to invest in public goods provision. In this case, a larger city council may have more partisan councilors aligned with the mayor, lowering the mayor's bargaining costs and increasing the equilibrium public goods provision.

I test these theoretical predictions by exploiting an exogenous variation in Brazilian city council sizes. In 2004, the judiciary reinterpreted the Constitution, unexpectedly changing the number of city councilors. Before the decision, towns could freely choose their legislature size, but in March 2004, the Superior Electoral Court created population thresholds, adding one councilor for each set of 47,619 inhabitants. Around these cutoffs, the decision represented an exogenous change in city council size, allowing me to investigate the effects of increasing the number of local legislators in Brazil.

Studying 2005 to 2008 Brazilian mayoral term, I find that increasing city council size changed the composition of the mayoral coalition and the number of appointed bureaucrats in the municipality. The extra city councilor had a 91 percent chance of belonging to the mayor's pre-electoral coalition. This enhanced chances of electing a councilor connected with the mayor decreased the bargaining costs for the mayor, freeing investments for public services. Moreover, the number of appointed bureaucrats increased by 104 with the additional councilor. In Brazil, more appointed bureaucrats lead to higher quality workers, better service provision, and more accountability \citep{toral2019benefits}.

More support for the mayors and more politically appointed bureaucrats correlate with city councilors' service provision expectations and legislation approved in Brazilian municipalities. I surveyed 174 former councilors who served during the 2005-2008 term, and they corroborate that mayors use bureaucratic appointments and granting favors to secure support. As the formal model shows, bargaining with councilors has a resource transfer component. By investigating what councilors prefer to provide to voters, I find that councilors demand job appointments, as these are helpful for their reelection chances and improve municipal welfare. These findings align with a novel dataset I build with 346,553 laws passed in 63 municipalities within ten thousand inhabitants away from the cutoffs. The legislation in municipalities on the left and on the right-hand side of the cutoffs comprised local public goods. However, municipalities with larger city councils had around 15.0 percent more public goods provision proposals, suggesting an increase in service provision legislation.

Finally, I show that the increase in legislature size significantly impacts municipal education and health care. Increasing the city council size by one legislator lowers infant mortality by 2.01 children per one thousand infants born and reduces postneonatal mortality by 0.90 children per one thousand infants that survived their first 28 days. A larger council size also increased enrollment by 2.58 children in elementary education classrooms without significantly diminish elementary school quality. These services are crucial for welfare and result from an eased bargaining between the mayor and the city council.

This paper contributes to three strands of the literature. First, this paper contributes to a large literature on multilateral bargaining. Repeated models of multilateral bargaining started with \citet{baron1989bargaining}, which extended the bilateral bargaining game proposed by \citet{rubinstein1982perfect} to more than two players. Several papers followed \citet{baron1989bargaining}, adding endogenous recognition power, changing the size of the required majority, adding partisanship, and testing the predictions of these models in the lab \citep{calvert2005legislative, choate2019partisan, agranov2020persistence, baranski2020determinants}.\footnote{See \citet{eraslan2019legislative} for a comprehensive review of the multilateral bargaining and \citet{ray2015coalition} for a review on coalition formation.} In this paper, I extend the \citet{baron1989bargaining} model under closed rules, adding the mayor's first move and political partisanship. Partisanship here is different from \citet{choate2019partisan} since I consider partisan as a zero-sum effect that is positive for politicians aligned with the mayor and negative otherwise. This paper investigates the often overlooked effect of legislature size on the bargaining equilibrium strategies.

Second, this paper contributes to the literature regarding the political and economic effects of larger legislatures. Following the contribution by \citet{Weingastetal1981}, the "law of 1/n" posits that larger legislatures increase the legislative pressure for providing targeted public goods to local supporters at the expense of the entire polity \citep{alptekin2021effect}. The corollary of the "law of 1/n" is that increasing legislature size should promote excessive public spending.\footnote{\citet{PrimoSnyder2008} show that legislature size may result in positive or negative expenditures, depending on factors such as the degree that local public goods can exclude non-locals or deadweight costs associated with taxation.} In the US metaphor, this would explain significant infrastructure spending in smaller districts. However, the "law of 1/n" literature left aside the public goods provision and welfare. This paper is one of the first papers to study the number of local legislators on welfare. The model provides a theoretical expectation based on the chances of electing a politician aligned with the mayor. The empirical estimates confirm the theoretical predictions, with the additional fact that legislature size has a null effect on expenditures in Brazil.

Finally, it contributes to a growing literature on municipal partisan politics. Most papers in this area estimate the effects of the mayor's party on political outcomes. For instance, \citet{gerber2011mayors} and \citet{de2016mayoral} find differences in governance when comparing Democratic and Republican mayors. Specifically about Brazil, \citet{gouvea2021partisanship} shows that social policy expenditure increase with left-wing mayors. \citet{frey2020larger}, using the discontinuities in city council size after the 2009 constitutional amendment that reverted the thresholds I study here, shows that changes in council size lower the support for candidates backed by the municipal mayors. This paper reinforces the view that partisanship matters locally, even in countries with weak ideological linkage.

@@

Several cities in the world are governed by the mayor-council system. In mayor-council cities, the power is divided between a mayor that controls the Executive branch and a city council that supports or sanctions the policies proposed by the mayor. In this setting, consider an strategic interaction between a mayor ($M$) and $N \geq 3$ city councilors, indexed in $i \in \{1, 2, \cdots, N \}$. The mayor has the prerogative of proposing a vector of policies that is voted by the city councilors. The provision of public goods is then a bargain between the mayor and the city council. Policy proposals are a combination of public goods provision $g$, rents for the mayor $r$, and a vector of transfer for city councilors \textbf{x}.

The city council votes the mayor's proposal, and if councilors accept it, the mayor implements it. Otherwise, a reversal policy takes place. Reversal policy in this context means that the mayor's proposal failed to secure the majority of council votes. The decision process leaves the mayor's office, and it is transferred to the council. The types of reversal policies are crucial for my argument, as it changes the relative strength of the council in their bargaining with the mayor. I investigate two types of reversal policies: a baseline reversal with no parties and a hybrid mechanism with partisan and non-partisan concerns. The differences between these two mechanisms are that the non-partisan councilors only care about rents, while partisan councilors care about both rents and policy.

To govern, the mayor must convince at least half of the city councilors to support her policy proposals. Similar to \citet{buenodemesquita2005logic}, I assume that the selectorate, i. e., the individuals that have a say in the mayor's policy proposals, are the $N$ city councilors. The winning coalition, the minimum number of members that must be convinced to implement the policy, is equal to half of the selectorate $N/2$. City councilors are moved by policy concerns $p$ and transfers received from the mayor or the councilor recognized as the proposer \textbf{x}. I leave the definition of these transfers open, as this facilitates the relationship with several instruments that mayors can use to pressure city councilors. Transfers may be granting electoral favors, favoring clientelistic brokers, spending in areas that the councilor has political or personal interests, providing portfolio (municipal offices) and public jobs within the municipal bureaucracy, or even campaign contributions and bribes. The use of transfers to affect the councilors' choices generate a bargaining cost of $C_G$. The bargaining costs vary with the reversal mechanism and the legislature size.

Finally, after learning the bargaining costs, the mayor decides a level of public goods provision ($g$) and pockets rents ($r$). Public goods provision helps the mayor to get reelected. Rents are for direct mayor's consumption, and they do not contribute to the mayor's electoral success.\footnote{This model, similarly to \citet{cowen1994rent} and \citet{buenodemesquita2005logic} combines public goods provision and rent extraction. For a dynamic bargaining model that incorporates transfers to districts and public goods provision, see \citet{battaglini2012legislative}.} This makes the mayor's expected utility a sum of the gains from rents and the benefits from reelection. Both the utilities from rents $u(.)$ and the probability of reelection $\pi(.)$ are concave functions, meaning that more rents or public goods increase the utility at a decreasing rate. The probability of reelection is multiplied by the benefit from holding office $B_M > 0$. This benefit captures the tangible and intangible gains that the mayor perceives from holding the public office. I assume that the benefits are high enough to rule out an equilibrium with only rents and transfers, which is equivalent to saying that mayors strongly prefer to be reelected. If the mayor's policy proposal is approved, then the expected utility for the mayor is:

@@

The policy choices of the mayor are subject to the municipal budget constraint. The municipality has $R > 0$ resources, and cannot run debts, meaning that the budget must be balanced. I also assume that there are enough resources for the mayor to govern.

Let $C_G$ be the bargaining costs. The budget balance constraint requires the offers that the mayor makes to the city councilors to satisfy the following inequality:

@@

The expected utility for the city councilors depends on the reversal policy. In the baseline non-partisan reversal, when the council rejects the mayor's proposal, a reversal stage starts with resources diminishing by a factor of $\delta \in (0,1)$ and the selection of one councilor randomly. The selected councilor becomes the new proposer. If her proposal is accepted, it is implemented. If councilors reject the offer again, the budget shrinks by a factor $\delta$, and another city councilor is recognized to propose. The process repeats until a proposal is finally accepted.

After studying the non-partisan baseline, I consider a situation that combines both partisan and non-partisan legislature concerns. Each councilor has a party affiliation. Party affiliations are mutually exclusive, and if a city councilor is aligned with the mayor's party, then she belongs to the government coalition $G \subset \{1, 2, \cdots, N \}$. Otherwise, the city councilor belongs to the opposition $O = \{1, 2, \cdots, N \} \setminus G$. The mayor's policy generates a political value of $p > 0$ for the councilors, and whether it adds or subtracts from the councilor's utility depends on the party affiliation of the councilor. If councilors reject the mayor's proposal, no public goods are provided ($p=0$). Moreover, in equilibrium, I assume that city councilors would prefer to receive transfers instead of providing public goods.\footnote{In the online supplemental materials, I consider a version of the model where public goods provision depends on the party affiliation of the proposer so that it affect choices in any potential history of the game. The equilibrium public goods provision level is even higher than in the model presented here, as city councilors prefer to approve the mayor's proposal than run the risk of an opposition member becoming the proposer.}

@@

\begin{enumerate}
  \item The mayor learns how many government $|G|$ and opposition $|O|$ legislators were elected.
  \item The mayor proposes a policy vector $(r, g, x)_M$.
  \item The city council votes the proposal.
  \begin{itemize}
    \item If the council accepts, the policy is implemented, and the game ends. 
    \item Otherwise, the reversal policy is implemented.
  \end{itemize}
  \item (Reversal Policy) The budget is discounted by a factor $\delta \in (0,1)$, and one councilor is randomly selected to make an offer. If the offer is accepted, it is implemented. If rejected, the reversal policy stage restart.\footnote{In section B.1 of the online supplemental materials, I solved three reversal mechanisms. The two in here, plus a mechanism with only political concerns.}
\end{enumerate}

The solution concept I use in this model is the Subgame Perfect Nash Equilibrium. A Subgame Perfect Nash Equilibrium requires that the strategies played are a Nash Equilibrium in each of the subgames of the game. The solution is found using backward induction. In the model with infinitely repeated proposals, any strategy can be shown to be a Nash equilibrium. Thus, a natural assumption is to extend the equilibrium concept to require the equilibrium to be stationary. A Stationary Subgame Perfect Equilibrium requires that, at each given point of the game, if a politician accepts an offer at time $k+1$, she should take the same offer at time $k$ \citep{rubinstein1982perfect}. Stationarity gives us an objective of finding a proposal that would be accepted at any stage in the game. Applying this reasoning, I can characterize a sequence of offers for each point $k$ in time. For the mayor, the optimal offer is at $k=0$, representing a no delay in policy implementation. Moreover, I assume that the players play no dominated strategies, even when there is less than sufficient support for implementing the policy.

For the solution to be of general interest, I solve bargaining costs for the expected number of opposition and government-aligned city councilors. This is because the solution for each fixed number of opposition and government councilors is trivial and does not help understand the overall relationship between council size and bargaining costs.

@@

Solving the game by backward induction requires us to start with the councilor's strategies. However, for exposure convenience, assume that I solved the game for the councilors and found the equilibrium bargaining costs $C_G(N)$ when there are $N$ city councilors. I now derive the optimal rents ($r$) and public goods provision ($g$) proposed by the mayor. The mayor benefits from public goods provision, as it increases the chances of her reelection. However, she prefers to invest as minimum as necessary in public goods and extract the remaining resources as political rents. The mayor's objective is to maximize her expected utility, subject to the municipal budget constraint.

@@

In equilibrium, the budget binds, and the first-order condition for an optimal public goods provision makes the marginal benefits to increase rents, in this case, marginal costs of providing the public goods, equal to the marginal benefits of reelection:

@@

All proofs are in section B.1 of the online supplemental materials. The proof's intuition involves finding what happens to the mayor's marginal utility when I increase both $g$ and $N$. As $N$ increases discretely on a complete lattice, I use monotone comparative statics to derive the conditions for increasing differences \citep{milgrom1994monotone, ashworth2006monotone}. Increasing differences mean that higher $N$ leads to higher equilibrium $g$.

Proposition \ref{prop1} provides the first empirical hypothesis to test: \textit{if the bargaining costs decrease when the legislature size increases, then the provision of the public goods increases when the legislature size increase}. Now I study each reversal mechanism to determine the condition under which the bargaining costs decrease when legislature size increases.

@@

I study the impact of legislature size on education and health care. I collect four indicators: two in education and two in health care.

>>>>

---
title:  |
        | Legislature Size and Welfare: Evidence from Brazil
author: 
  - Gabriel Cepaluni[^gabriel]
  - Danilo Freire[^danilo]
  - Umberto Mignozzetti[^umberto]
date: \today
thanks: "We thank Cedric Antunes, Victor Araujo, Eric Arias, Taylor Boas, Natalia Bueno, Andrea Freitas, Saad Gulzar, James Mahon, Lucas Novaes, Nelson Ruiz, Catarina Roman, David Skarbek, and Denis Stukal for their valuable comments. We also thank the participants of seminars and conferences at CEPESP, CUNY, INSPER, IPSA, LACEA, LASA, MPSA, Notre Dame, NYU, PUC Chile, SPSA, and Uppsala. We gratefully acknowledge the financial support of the São Paulo Research Foundation (FAPESP), grant number 2017/07290-5. All materials required to replicate this study are available at <https://github.com/umbertomig/legislature-size-welfare>."
abstract: "How does legislature size impact public service provision? Despite the importance of institutional design for democratic governance, the effect of particular legislative features on citizen welfare remains little understood. In this paper, we use a formal model to show that increasing legislature size improves the provision of public goods. We argue that changes in bargaining costs depend on whether additional legislators share the Executive's party affiliation: More opposition members reduce the Executive's public goods provision, while more government-aligned members increase it. We test this theory by exploiting sharp discontinuities in city council size in Brazil. We show that an additional city councilor has a 91% chance of belonging to the mayoral coalition, and this significantly improves primary school enrollment and infant mortality measures. To explore possible mechanisms, we surveyed former councilors and analyzed 346,553 bills proposed between 2005 and 2008. This paper has implications for the design of representative institutions."
abstractspacing: double
keywords: Legislature Size; Legislative Bargaining; Local Public Goods Provision; Regression Discontinuity Design; Brazil
jelcodes: H41; H80; I31; I38
fontsize: 11pt
margin: 2cm
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
    md_extensions: +raw_attribute
---

@@

@@

Legislative institutions are crucial for citizen welfare and service provision in democratic countries. Historically, the division of power and the birth of a representative government placed legislatures at the center of the political stage. In modern democracies, most countries have assemblies at all levels of government. These institutions regulate the quantity and quality of legislation, check and veto other political powers, and influence policymaking. Moreover, legislatures also determine taxation, government spending and oversee public programs \citep{auriol2012optimal, north1989constitutions, poulsen2018political, stasavage2011states, Weingastetal1981, weingast1988industrial}.

Regardless of eventual differences in electoral rules, social composition, and internal procedures, all legislatures share one feature: they are collective bodies. As such, the number of representatives in parliament has important consequences for whole polity. Indeed, legislature size is known to have a significant effect on collective action strategies, the bargaining power between state branches, government spending, and representation \citep{allenrepresentation, crain1979cost, freire2021legislature, PrimoSnyder2008, rogers2002free, PrimoSnyder2008, Weingastetal1981}. However, how legislature size impacts social welfare remain less understood. On the one hand, enlarged legislatures may increase government spending and lawmaking, which can lead to more public goods \citep{Weingastetal1981}. On the other hand, adding a legislator may increase the number of veto players and exacerbate collective action problems, thus hindering service provision \citep{gottlieb2019countervailing, tsebelis2002veto}. Therefore, what is the effect of larger legislatures on public service provision?

In this paper, we propose a local legislature size and welfare theory that incorporates two crucial features: Executive leadership and party politics. First, in mayor-council municipalities, mayors tend to be central in local public service decisions. Mayors have the prerogative to boost public investments, like presidents in federal governments. Second, partisanship is essential at the local level, even in settings where the ideological attachment is weak, such as in most young democracies. Party attachment structures political views \citep{samuels2014power}, and local-level elites usually belong to different factions and parties. They fight for power in the municipality by controlling various parties and forming distinct pre-electoral coalitions.\footnote{Elite political competition in the local level has also been linked to violence and conflict in Brazil and elsewhere \citep{hoelscher2015politics, mignozzetti2018predatory}.} In many cities, family ties may be substitutes for the ideological component, but influential local leaders usually sort themselves into different political groups \citep{ferraz2020political, nunes1977coronelismo}.

This paper incorporates partisanship and mayoral first-move advantage to a \citet{baron1989bargaining} bargaining model. Mayors make the initial policy proposal that consists of an investment in public goods provision, a decision to extract rents for personal gains, and a vector of side-transfers for councilors. If they fail to secure majoritarian support from councilors, a reversal stage occurs, where the decision leaves the mayor's office and councilors propose the public investments. In this mayor-council bargaining process, costs come from the side-transfers required to secure a majority among councilors. We show that in a model without partisan concerns, bargaining costs always increase with larger city councils, which in turn lower public goods provision and rents. However, partisanship decreases bargaining costs when the chances of electing councilors aligned with the mayor are sufficiently high. Lower bargaining costs release funds for mayors to invest in public goods provision. In this case, a larger city council may have more partisan councilors aligned with the mayor, reducing the mayor's bargaining costs and increasing the equilibrium public goods provision.

We test these theoretical predictions by exploiting an exogenous variation in Brazilian city council sizes. In 2004, the judiciary reinterpreted the Brazilian Constitution and unexpectedly changed the number of councilors in several municipalities. Until then, towns could freely determine their legislature size, but in March 2004 the Supreme Electoral Court created population thresholds and added one councilor to each set of 47,619 inhabitants. Around these cutoffs, the decision represented an exogenous change in city council size, allowing us to investigate the effects of increasing the number of local legislators in Brazil.

We find that larger city councils changed the composition of mayoral coalitions and the number of appointed bureaucrats in the municipality. The extra city councilor had a 91 percent chance of belonging to the mayor's pre-electoral coalition. This enhanced the chances of electing a councilor connected with the mayor, which then decreased the bargaining costs for the mayor and boosted investments in public services. Moreover, the number of appointed bureaucrats increased by 104 with the additional councilor. In Brazil, appointing bureaucrats leads to higher quality workers, better service provision, and more accountability \citep{toral2019benefits}.

Higher mayor support and more appointed bureaucrats correlate with city councilors' service provision expectations and legislation approved in Brazilian municipalities. We surveyed 174 former councilors who served during the 2005-2008 term, and they corroborate that mayors use bureaucratic appointments and political favors to secure support. As the formal model shows, bargaining with councilors requires offering transfers to the representatives. By investigating what councilors prefer to provide to voters, we find that councilors mainly demand job appointments, as these increase their reelection chances and improve municipal welfare. These findings align with a novel dataset we built with 346,553 laws passed in 63 municipalities within ten thousand inhabitants away from the cutoffs. The legislation in municipalities on the left and on the right-hand side of the cutoffs mentioned local public goods. However, municipalities with larger city councils had around 15 percent more public goods provision proposals, what is in line with our expectations.

Finally, we show that the increase in legislature size significantly impacts municipal education and health care. Increasing the city council size by one legislator lowers infant mortality by 2.01 children per one thousand infants born and reduces postneonatal mortality by 0.90 children per one thousand infants that survived their first 28 days. A larger council size also increased enrollment by 2.58 children in elementary education classrooms without significantly diminish elementary school quality. These services are crucial for welfare and result from an eased bargaining between the mayor and the city council.

This paper contributes to three strands of the literature. First, this paper dialogues with a large literature on multilateral bargaining. Repeated models of multilateral bargaining started with \citet{baron1989bargaining}, who extended the bilateral bargaining game proposed by \citet{rubinstein1982perfect} to more than two players. Several papers followed \citet{baron1989bargaining}, manipulating endogenous recognition power, the size of the required majority, and partisanship in the models \citep{agranov2020persistence, baranski2020determinants, calvert2005legislative, choate2019partisan}.\footnote{See \citet{eraslan2019legislative} for a comprehensive review of multilateral bargaining and \citet{ray2015coalition} for a review on coalition formation.} In this paper, we extend the \citet{baron1989bargaining} model under closed rules and add the mayor's first move and political partisanship. Partisanship here is different from \citet{choate2019partisan} since we consider partisanship as a zero-sum effect that is positive for politicians aligned with the mayor and negative otherwise. This paper investigates the often overlooked effect of legislature size on the bargaining equilibrium strategies.

Second, this paper contributes to the literature on the political and economic effects of larger legislatures. Following the contribution by \citet{Weingastetal1981}, the "law of $1/n$" posits that larger legislatures increase the legislative pressure for providing targeted public goods to local supporters at the expense of the entire polity \citep{freire2021legislature}. The corollary of the "law of $1/n$" is that increasing legislature size promotes excessive public spending.\footnote{\citet{PrimoSnyder2008} show that legislature size may result in positive or negative expenditures, depending on factors such as the degree that local public goods can exclude non-locals or deadweight costs associated with taxation.} In the US metaphor, this would explain significant infrastructure spending in smaller districts. However, the "law of $1/n$" literature rarely mentions the relationship between public goods provision and welfare. This paper is one of the first to study how the number of local legislators affects welfare. Our model provides a theoretical expectation based on the chances of electing a politician aligned with the mayor. The empirical estimates confirm our theoretical predictions, with the additional fact that legislature size has a null effect on expenditures in Brazil.

Finally, our article also contributes to a growing literature on municipal partisan politics. Most papers in this area estimate the effects of the mayor's party on political outcomes. For instance, \citet{gerber2011mayors} and \citet{de2016mayoral} find differences in governance when comparing Democratic and Republican mayors. Specifically about Brazil, \citet{gouvea2021partisanship} shows that social policy expenditure increase with left-wing mayors. \citet{frey2020larger}, using discontinuities in city council size after the 2009 constitutional amendment that reverted the thresholds we study here, shows that changes in council size lower the support for candidates backed by the municipal mayors. This paper reinforces the view that partisanship matters locally, even in countries with weak ideological linkage.

@@

Several cities in the world are governed by the mayor-council system. In mayor-council cities, the power is divided between a mayor that controls the Executive branch and a city council that supports or sanctions the policies proposed by the mayor. In this setting, consider an strategic interaction between a mayor ($M$) and $N \geq 3$ city councilors indexed in $i \in \{1, 2, \cdots, N \}$. The mayor has the prerogative of proposing a vector of policies that is voted by the city councilors. The provision of public goods is then a bargain between the mayor and the city council. Policy proposals are a combination of public goods provision $g$, rents for the mayor $r$, and a vector of transfer for city councilors \textbf{x}.

The city council votes the mayor's proposal, and if councilors accept it, the proposal is implemented. Otherwise, a reversal policy takes place. Reversal policy in this context means that the mayor's proposal failed to secure the majority of council votes. The decision process then leaves the mayor's office and it is transferred to the council. The types of reversal policies are crucial for our argument, as it changes the relative strength of the council when they bargain with the mayor. We investigate two types of reversal policies: a baseline reversal with no parties and a hybrid mechanism with partisan and non-partisan concerns. The differences between these two mechanisms are that the non-partisan councilors only care about rents, while partisan councilors care about both rents and policy.

The mayor must convince at least half of the city councilors to support her policy proposals. Similar to \citet{buenodemesquita2005logic}, we assume that the selectorate, that is, the individuals who have a say in the mayor's policy proposals, are the $N$ city councilors. The winning coalition, the minimum number of members that must be convinced to implement the policy, is equal to half of the selectorate ($N/2$). City councilors are motivated by policy concerns $p$ and transfers received from the mayor or the councilor recognized as the proposer \textbf{x}. We leave the definition of these transfers open, as this allows us to analyze how mayors combine transfers with other incentives to pressure city councilors. Exemples of transfers are granting electoral favors, favoring clientelistic brokers, spending in areas that the councilor has political or personal interests, providing portfolio (municipal offices) and public jobs within the municipal bureaucracy, or even giving campaign contributions and bribes. The use of transfers to influence the councilors' choices generate a bargaining cost of $C_G$. Bargaining costs vary according to the reversal mechanism and the size of the legislature.

Finally, after learning the bargaining costs, the mayor provides a level of public goods provision $g$ and pockets rents $r$. Public goods provision helps the mayor to get reelected. Rents are for direct mayor's consumption, and they do not contribute to the mayor's electoral success.\footnote{This model, similar to \citet{cowen1994rent} and \citet{buenodemesquita2005logic}, combines public goods provision and rent extraction. For a dynamic bargaining model that incorporates transfers to districts and public goods provision, see \citet{battaglini2012legislative}.} This makes the mayor's expected utility a sum of the gains from rents and the benefits from reelection. Both the utilities from rents $u(.)$ and the probability of reelection $\pi(.)$ are concave functions, meaning that more rents or public goods increase the utility at a decreasing rate. The probability of reelection is multiplied by the benefit from holding office $B_M > 0$. This benefit captures the tangible and intangible gains that the mayor receives from holding the public office. We assume that the benefits are high enough to rule out an equilibrium with only rents and transfers, which is equivalent to saying that mayors strongly prefer to be reelected. If the mayor's policy proposal is approved, then the expected utility for the mayor is:

@@

The policy choices of the mayor are subject to municipal budget constraints. The municipality has $R > 0$ resources and cannot run debts, meaning that the budget must be balanced. We also assume that there are enough resources for the mayor to govern.

Let $C_G$ be the bargaining costs. The budget balance constraint requires that the offers the mayor makes to the city councilors must satisfy the following inequality:

@@

The expected utility for the city councilors depends on the reversal policy. In the baseline non-partisan reversal, when the council rejects the mayor's proposal, a reversal stage starts with resources diminishing by a factor of $\delta \in (0,1)$ and the selection of one councilor randomly. The selected councilor becomes the new proposer. If her proposal is accepted, it is implemented. If councilors reject the offer again, the budget shrinks by $\delta$, and another city councilor is recognized to propose. The process repeats until a proposal is finally accepted.

We now consider a situation that combines both partisan and non-partisan legislature concerns. Each councilor has a party affiliation. Party affiliations are mutually exclusive, and if a city councilor is aligned with the mayor's party, then she belongs to the government coalition $G \subset \{1, 2, \cdots, N \}$. Otherwise, the city councilor belongs to the opposition $O = \{1, 2, \cdots, N \} \setminus G$. The mayor's policy generates a political value of $p > 0$ for the councilors, and whether it adds or subtracts from the councilor's utility depends on the party affiliation of the councilor. If councilors reject the mayor's proposal, no public goods are provided ($p=0$). Moreover, in equilibrium, we assume that city councilors would prefer to receive transfers instead of providing public goods.\footnote{In the Supplementary Materials, we consider a version of the model where public goods provision depends on the party affiliation of the proposer so that it affects choices in any potential history of the game. The equilibrium public goods provision level is even higher than in the model presented here, as city councilors prefer to approve the mayor's proposal than run the risk of an opposition member becoming the proposer.}

@@

\begin{enumerate}
  \item The mayor learns how many government $|G|$ and opposition $|O|$ legislators were elected.
  \item The mayor proposes a policy vector $(r, g, x)_M$.
  \item The city council votes the proposal.
  \begin{itemize}
    \item If the council accepts the proposal, the policy is implemented and the game ends. 
    \item Otherwise, the reversal policy is implemented.
  \end{itemize}
  \item (Reversal Policy) The budget is discounted by a factor $\delta \in (0,1)$, and one councilor is randomly selected to make an offer. If the offer is accepted, it is implemented. If rejected, the reversal policy stage restart.\footnote{In section A.1 of the Supplementary Materials, we solve three reversal mechanisms. The two we present here, plus a mechanism with only political concerns.}
\end{enumerate}

The solution concept we use in this model is the subgame perfect Nash equilibrium. This game requires that the strategies follow a Nash equilibrium in each subgame. We find the solution using backward induction. In the model with infinitely repeated proposals, any strategy can be shown to be a Nash equilibrium. Thus, a natural assumption is to extend the equilibrium concept to require the equilibrium to be stationary. A stationary subgame perfect equilibrium requires that, at each given point of the game, if a politician accepts an offer at time $k+1$, she should take the same offer at time $k$ \citep{rubinstein1982perfect}. Stationarity gives us a method to find a proposal that would be accepted at any stage of the game. Hence, we can characterize a sequence of offers for each point $k$ in time. For the mayor, the optimal offer is at $k=0$, representing a no delay in policy implementation. Moreover, we assume that the players play no dominated strategies, even when there is insufficient support for the policy.

For the solution to be of general interest, we solve the bargaining costs for the expected number of opposition and government-aligned city councilors. This is because the solution for each fixed number of opposition and government councilors does not help us understand the overall relationship between council size and bargaining costs.

@@

Solving the game by backward induction requires us to start with the councilors' strategies. However, to conserve space, assume that we solved the game for the councilors and found the equilibrium bargaining costs $C_G(N)$ when there are $N$ city councilors. We now derive the optimal rents ($r$) and public goods provision ($g$) proposed by the mayor. The mayor benefits from public goods provision, as it increases the chances of her reelection. However, she prefers to invest as minimum as necessary in public goods and extract the remaining resources as political rents. The mayor's objective is to maximize her expected utility, subject to the municipal budget constraint.

@@

In equilibrium, the budget binds, and the first-order condition for an optimal public goods provision makes the marginal benefits to increase rents. In this case, it means that the marginal costs of providing the public goods is equal to the marginal benefits of reelection:

@@

All proofs are in section A.1 of the Supplementary Materials. The proof's intuition involves finding what happens to the mayor's marginal utility when we increase both $g$ and $N$. As $N$ increases discretely on a complete lattice, we use monotone comparative statics to derive the conditions for increasing differences \citep{milgrom1994monotone, ashworth2006monotone}. Increasing differences mean that higher $N$ leads to higher equilibrium $g$.

Proposition \ref{prop1} provides the first empirical hypothesis to test: \textit{if the bargaining costs decrease when the legislature size increases, then the provision of the public goods increases when the legislature size increase}. Now we study each reversal mechanism to determine the condition under which the bargaining costs decrease when legislature size increases.

@@

We now study the impact of legislature size on education and health care. We use four indicators in this paper: two in education and two in health care.
